Forecasting the age structure of the scientific workforce in Australia

### 2025-09-09 age_structure_forecasts.qmd
Minor improvements

```{r setup, include=FALSE}
library(targets)

@@

The economic implications of workforce age structure shifts are well-documented [e.g., @Bloom2007], affecting productivity, pensions and superannuation, and skill shortages [@PC2013; @OECD2019a; @OECD2019b; @pension]. The social implications are also significant, with an aging workforce leading to changes in workplace dynamics, potential problems with intergenerational knowledge transfer, and the need for policies that support older workers. Yet this problem does not appear to have been previously addressed from a statistical modeling perspective.

@@

* $D_{i,x,t} \sim \text{Binomial}(P_{i,x,t}, q_{i,x,t})$, with $q_{i,x,t}$ being the probability of death for group $i$ at age $x$ in year $t$;
* $R_{i,x,t} \sim \text{Binomial}((P_{i,x,t}-D_{i,x,t}), r_{i,x,t})$, with $r_{i,x,t}$ being the probability of retirement from group $i$ at age $x$ in year $t$.

@@

To obtain forecasts of $P$, we use functional time series models [@HU07] for each of $q$, $G$, and $E$, and simulate future sample paths.

@@

This analysis focuses on the Natural and Physical Sciences as defined in the Australian Standard Classification of Education (ASCED) by the @ABSASCED. We refer to ASCED's Narrow Fields as "disciplines"; these comprise Physics and Astronomy, Mathematical Sciences, Chemical Sciences, Earth Sciences, Biological Sciences, Other Natural and Physical Sciences, and Natural and Physical Sciences not further defined (n.f.d.). @tbl-disciplines lists the detailed fields within each scientific discipline.

@@

The demographic growth-balance equation (@eq-dgb), when rearranged, provides an expression for the remainder including net migration and career changes:
$$
E_{i,x,t} = P_{i,x+1,t+1} - P_{i,x,t} - D_{i,x,t} - R_{i,x,t} - g_xG_{i,t},
$$ {#eq-remainder}
However, we do not have data on $D_{i,x,t}$ and $R_{i,x,t}$, so we replace these by their expected values, $P_{i,x,t} q_{x,t}$ and $P_{i,x,t}(1-q_{x,t}) r_x$, respectively. We can only estimate remainders up to 2020 because we need data for both year $t$ and year $t+1$ in @eq-remainder, and our working population data only extends to 2021. The estimated remainders are shown in @fig-remainder.

@@

To forecast future working population numbers, $P_{i,x,t}$, we use the demographic growth-balance equation (@eq-dgb), with forecasts of the components $G_{i,t}$, $q_{x,t}$, and $E_{i,x,t}$. To account for the combined uncertainty from all of the forecast components, we use a simulation-based approach, running 1000 simulations to generate a forecast distribution of possible future outcomes.

@@

```{r}
#| label: fig-grad-forecasts
#| fig-cap: "Forecast of $G_{i,t}$: the number of graduates by discipline, 2024-2035, with 90% prediction intervals, based on historical data from 2006--2023."
tar_read(fig_grad_forecasts)
```

@@

```{r}
#| label: fig-death-probs-forecasts
#| fig-cap: "Forecasts of $q_{x,t}$: age-specific probabilities of death (on a logarithmic scale) for 2035, with 90% prediction intervals, based on historical data from 1971--2021."
tar_read(fig_model_mxt3)
```

@@

```{r}
#| label: fig-net-migration-disciplines
#| fig-cap: "Forecasts of $E_{i,x,t}$: the remainder by discipline for 2035, with 90% prediction intervals, based on historical data from 2006--2020."
tar_read(fig_model_Ext3)
```

@@

We use the demographic growth-balance model @eq-dgb to simulate future populations, using the models described above for graduate completions, death probabilities and the remander terms. The following steps outline the process:

A total of 1000 simulations are run to obtain a distribution of possible future population scenarios.
The average of the 1000 simulations provides the mean age-specific forecast, while quantiles estimate forecast uncertainty. @fig-forecast-working-population presents the mean and 90% prediction intervals for 2025 and 2035.

```{r}
#| label: fig-forecast-working-population
#| fig-cap: "Projected age distribution of the working population in 2025 and 2035 across scientific disciplines. The shaded regions represent the 90% prediction intervals, and the solid lines indicate the mean estimates for each year."
#| fig-height: 5.5
# 2025 and 2035
#tar_read(fig23)
```

@@

Between 2025 and 2035, the forecasts indicate an aging workforce, with increases in the number of scientists in their late 30s, and from ages 45 to 80. This trend is particularly evident in the wider gap at ages 45-60 between the forecasted distributions over the ten-year period.

>>>>

```{r setup, include=FALSE}
library(targets)
library(ggplot2)

@@

The economic implications of workforce age structure shifts are well-documented [e.g., @Bloom2007], affecting productivity, pensions and superannuation, and skill shortages [@PC2013; @OECD2019a; @OECD2019b; @pension]. The social implications are also significant, with an aging workforce leading to changes in workplace dynamics, potential problems with intergenerational knowledge transfer, and the need for policies that support older workers. Yet this problem does not appear to have been previously addressed from a statistical modelling perspective.

@@

* $D_{i,x,t} \sim \text{Binomial}(P_{i,x,t}, q_{i,x,t})$, with $q_{i,x,t}$ being the probability of death for group $i$ at age $x$ in year $t$; and
* $R_{i,x,t} \sim \text{Binomial}((P_{i,x,t}-D_{i,x,t}), r_{i,x,t})$, with $r_{i,x,t}$ being the probability of retirement from group $i$ at age $x$ in year $t$.

@@

This leads to the simpler model
$$
P_{i,x+1,t+1} = P_{i,x,t} - D_{i,x,t} - R_{i,x,t} + g_xG_{i,t} + E_{i,x,t},
$$ {#eq-dgb2}
where

* $D_{i,x,t} \sim \text{Binomial}(P_{i,x,t}, q_{x,t})$; and
* $R_{i,x,t} \sim \text{Binomial}((P_{i,x,t}-D_{i,x,t}), r_{x})$.

We use functional time series models [@HU07] for $q_{x,t}$ and $E_{i,x,t}$, a global ARIMA models for $G_{i,t}$, and univariate ARIMA models for the principal component scores of the functional time series models.

To forecast future working population numbers, $P_{i,x,t}$, $t>T$, we simulate future sample paths of each of the components $G_{i,t}$, $q_{x,t}$, and $E_{i,x,t}$, simulate $D_{i,x,t}$ and $R_{i,x,t}$ from their respective Binomial distributions, and then use the demographic growth-balance equation @eq-dgb2 iteratively to obtain $P_{i,x,t}$ for $t=T+1,T+2,\dots$. This simulation-based approach allows us to capture the uncertainty in each of the components, leading to a distribution of possible future outcomes for $P_{i,x,t}$.

@@

To illustrate the methodology, we consider the Natural and Physical Sciences as defined in the Australian Standard Classification of Education (ASCED) by the @ABSASCED. We refer to ASCED's Narrow Fields as "disciplines"; these comprise Physics and Astronomy, Mathematical Sciences, Chemical Sciences, Earth Sciences, Biological Sciences, Other Natural and Physical Sciences, and Natural and Physical Sciences not further defined (n.f.d.). @tbl-disciplines lists the detailed fields within each scientific discipline.

@@

The demographic growth-balance equation (@eq-dgb2), when rearranged, provides an expression for the remainder including net migration and career changes:
$$
E_{i,x,t} = P_{i,x+1,t+1} - P_{i,x,t} - D_{i,x,t} - R_{i,x,t} - g_xG_{i,t},
$$ {#eq-remainder}
However, we do not have data on $D_{i,x,t}$ and $R_{i,x,t}$, so we replace these by their expected values, $P_{i,x,t} q_{x,t}$ and $P_{i,x,t}(1-q_{x,t}) r_x$, respectively. We can only estimate remainders up to 2020 because we need data for both year $t$ and year $t+1$ in @eq-remainder, and our working population data only extends to 2021. The estimated remainders are shown in @fig-remainder.

@@

@@

```{r}
#| label: fig-grad-forecasts
#| fig-cap: "Forecast of $G_{i,t}$: the number of graduates by discipline, 2024-2035, based on historical data from 2006--2023. The shaded regions represent the 90% prediction intervals, and the solid lines indicate the mean estimates."
tar_read(fig_grad_forecasts)
```

@@

```{r}
#| label: fig-death-probs-forecasts
#| fig-cap: "Forecasts of $q_{x,t}$: age-specific probabilities of death (on a logarithmic scale) for 2035, based on historical data from 1971--2021. The shaded regions represent the 90% prediction intervals, and the solid lines indicate the mean estimate."
tar_read(fig_model_mxt3)
```

@@

```{r}
#| label: fig-net-migration-disciplines
#| fig-cap: "Forecasts of $E_{i,x,t}$: the remainder by discipline for 2035, based on historical data from 2006--2020. The shaded regions represent the 90% prediction intervals, and the solid lines indicate the mean estimates."
tar_read(fig_model_Ext3)
```

@@

We use the demographic growth-balance model @eq-dgb2 to simulate future populations, using the models described above for the components. The following steps outline the process:

A total of 1000 simulations are run to obtain a distribution of possible future population scenarios.
The average of the 1000 simulations provides the mean age-specific forecast, while quantiles estimate forecast uncertainty. @fig-forecast-working-population presents the mean and 90% prediction intervals for 2035.

```{r}
#| label: fig-forecast-working-population
#| fig-cap: "Forecasts of $P_{i,x,t}$: the working population by discipline for 2035. The shaded regions represent the 90% prediction intervals, and the solid lines indicate the mean estimates."
#| fig-height: 5.5
# 2025 and 2035
tar_load(fig_Pxt_future_discipline)
i <- length(fig_Pxt_future_discipline)
fig_Pxt_future_discipline[[i]] +
  guides(fill = "none") +
  labs(title = "Forecasts of working population by discipline: 2035")
```

@@

Between 2025 and 2035, the forecasts indicate an aging workforce, with increases in the number of scientists in their late 30s, and from ages 45 to 80. This trend is particularly evident in the wider gap at ages 45--60 between the forecasted distributions over the ten-year period.
